Chemical potential of liquids and mixtures via Adaptive Resolution Simulation

### 2014-05-08 writeDown/pub.jcp/chem-pot-jcp.tex
my revision

\def\pd#1#2{\frac{\partial #1}{\partial #2}}
\renewcommand{\vec}[1]{\mbox{\boldmath \small $#1$}}
\newcommand{\concen}{C}
\setlength\parindent{2em} 
\newcommand{\recheck}[1]{{\color{red} #1}}
\newcommand{\redc}[1]{{\color{red} #1}}            
\newcommand{\bluec}[1]{{\color{blue} #1}}            
\newcommand{\vect}[1]{\textbf{\textit{#1}}}
\newcommand{\dof}{{\textrm{DOF}}}
\newcommand{\AT}{{\textrm{{AT}}}}
\newcommand{\EX}{{\textrm{EX}}}
\newcommand{\CG}{{\textrm{CG}}}
\newcommand{\HY}{{\Delta}}
\newcommand{\rdf}{{\textrm{rdf}}}
\newcommand{\thf}{{\textrm{th}}}
\newcommand{\res}{{\textrm{rep}}}
\newcommand{\ext}{{\textrm{extra}}}
\newcommand{\exc}{{\textrm{exc}}}
\newcommand{\thermo}{{\textrm{Q}}}
\newcommand{\hadress}{{\textrm{H}}}
\newcommand{\dadress}{{\textrm{D}}}

@@

\bluec{
  Unified of terminology:
  \begin{itemize}
  \item interface region $\rightarrow$ transition region
  \item hybrid region $\rightarrow$ transition region
  \end{itemize}
}

\section{Introduction}
The chemical potential represents an important thermodynamic information for any system, in particular for liquids, where the possibility of combining different substances for forming optimal mixtures is strictly related to knowledge of the chemical potential of each component in the mixture environment. In this perspective, molecular simulation represents a powerful tool for predicting the chemical potential of complex molecular systems. Popular, well established methodologies in Molecular Dynamics (MD) are Widom particle insertion (IPM) \cite{widom} and thermodynamic integration (TI) \cite{ti}. IPM is computationally very demanding often beyond a reasonable limit even in presence of large computational resources, but upon convergence, is rather accurate. TI is computationally convenient but specifically designed to calculate the chemical potential and thus  it may not be optimal for employing MD for studying other properties. In fact TI requires artificial modification of the atomistic interactions (see Appendix).
Recently we have suggested that the chemical potential could be calculated by employing the Adaptive Resolution Simulation method in its Grand Canonical-like formulation (GC-AdResS) \cite{prl12,jctchan,prx}.
AdResS was originally designed  to interface regions of space at different levels of molecular resolution within one simulation set up. This allows for large and efficient multiscale simulations where the high resolution region is restricted to a small portion of space and the rest of the system is at coarser level. The recent version of the method, GC-AdResS, given its theoretical framework, should automatically calculate the chemical potential during the process of initial equilibration: in this work we prove that this is indeed the case and report results for the chemical potential for various liquids and mixtures of particular relevance in (bio)-chemistry and material science. We compare our results with those from full atomistic TI and find a satisfactory agreement. This agreement allows us to conclude that every time a multiscale GC-AdResS is performed, $\mu^{ex}$ is automatically calculated for each liquid component and implicitly confirm that the basic thermodynamics of the system is well described by the method.
\recheck{Moreover, in recent work AdResS has been merged with the MARTINI force field \cite{matej-sw1,matej-sw2}. In this context, the possibility of having a quantity like the chemical potential under \bluec{what is ``explicit control''} can be used as a further argument for the validity of the method in applications to large systems of biological interest.}
Below we provide the basic technical ingredients of GC-AdResS which are relevant for the calculation of the chemical potential, more specific details can be found in \cite{jctchan, prx}.

@@

from the corresponding coarse-grained potential, $\vect r$ is the center of mass (COM) position of

@@

to $1$ (or vice versa) in the interface region ($\Delta$) where the lower resolution is then

@@

as illustrated in Fig.\ref{fig1}. .  
\item In the interface region a thermodynamic force acting on the center of mass (CM) of each molecule and a locally acting thermostat are added to assure the overall thermodynamic equilibrium at a given temperature. The thermodynamic force is defined in such a way that $p_{AT}+\rho_{0}\int_{\Delta}{\bf F}_{th}({\bf r})d{\bf r}=p_{CG}$, where $p_{AT}$  is the target pressure of the atomistic system (region), $p_{CG}$ is the pressure of the coarse-grained model, $\rho_{0}$ is the target molecular density of the atomistic system (region)\cite{prl12}. An additional locally acting thermostat is added to take care of the loss/gain of energy in the transition region.

@@

$w(x)$ (in green) is defined, and the region on the left, is the high resolution region (atomistic). It must be noticed that differently from the original AdResS, in GC-AdResS the range of definition of $w(x)$ is extended of an amount of ${\bf R}_{c}$. The extension of this additional region is equal to the cutoff radius of the atomistic interactions and $w(x)$ takes the constant value of $1$. The consequence is that molecules in the atomistic region interact with the rest of the system always and only via well defined atomistic interactions. This characteristic, in turn, allows to write an exact Hamiltonian for the atomistic region and thus treat the system in a Grand-Canonical fashion.\label{fig1}}

@@

In \cite{prx} we have defined necessary conditions in $\Delta$ such that the spatial probability distribution of the full-atomistic reference system was reproduced up to a certain (desired) order in the atomistic region of the adaptive system. \recheck{We have defined the $i$th order of a spatial (configurational) probability distribution of $N$ molecules, $p({\bf r}_{1}, \cdots, {\bf r}_{N})$, as:

@@

p^{(i)}({\bf r}_{1}, \cdots, {\bf r}_{i})=\int p({\bf r}_{1}, \cdots, {\bf r}_{i},{\bf r}_{i+1}, \cdots, {\bf r}_{N}) \: d{\bf r}_{i+1}\cdots d{\bf r}_{N}

@@

The first order, often mentioned in this work corresponds to the molecular number density $\rho({\bf r})$.}

@@

 In the simulation set up, ${\bf F}_{th}$ is calculated via an iterative procedure as a function of the molecular density in $\Delta$. The iterative scheme consists of calculating ${\bf F}^{i+1}_{th}({\bf r})=\frac{1}{\rho_{0}\kappa_{T}}\nabla \rho^{i}({\bf r})$, ($\kappa_{T}$ the isothermal compressibility) and the force is considered converged when in $\Delta$ the target density $\rho_{0}$ is reached. As a result, ${\bf F}_{th}({\bf r})$, acting in $\Delta$, assures that there are no artificial density variations across the system, thus it allows to accurately reproduce the first order of the probability distribution in the atomistic region. Higher orders can be systematically achieved by imposing in $\Delta$ a corrective force proportional to the gradient of the corresponding higher order of the probability distribution; for example, the gradient of the CM-CM radial distribution function for the second order. Next it was proved that indeed the target Grand Canonical distribution, that is the probability distribution of a subsystem (of the size of the atomistic region in GC-AdResS) in a large full atomistic simulation is accurately reproduced. A large number of tests were performed and the reproduction by GC-AdResS of the probability distribution was numerically proved up to (at least) the third order, more than sufficient in MD simulations. Within this framework it was finally shown that the sum of work of ${\bf F}_{th}({\bf r})$ and that of the thermostat corresponds to the difference in chemical potential between the atomistic and coarse-grained resolution; this subject is treated in the next section.
\section{Calculation of $\mu$}
In Ref.\cite{prx} it has been shown that the chemical potential of the atomistic and coarse-grained resolution are related by the following formula: 
\mu_{CG}=\mu_{AT}+\omega_{th}+\omega_{Q}
\label{mu}
 with $\mu_{CG}$ the chemical potential of the coarse-grained system (in GC-AdResS this corresponds to a liquid of generic spheres), $\mu_{AT}$  the chemical potential of the atomistic system, $\omega_{th}=\int_{\Delta}{\bf F}_{th}({\bf r})d{\bf r}$ the work of the thermodynamic force in the transition region, $\omega_{Q}$ the work/heat provided by the thermostat in order to slowly equilibrate the inserted/removed degrees of freedom in the transition region. $\omega_{Q}$ is composed by two parts, one, called $\omega_{extra}$, which compensates the dissipation of energy due to the change of interactions in $\Delta$, and another, $\omega_{DOF}$, which is related to the equilibration of the reinserted/removed degrees of freedom (rotational and vibrational). \recheck{ While the determination of $\omega_{DOF}$ is not required for the our final aim (that is the calcualtion of the excess chemical potential, as explain later on), the calculation of $\omega_{Q}$ is very relevant. However this calculation is not straightforward and we have proposed to introduce an auxiliary Hamiltonian approach where the coarse-grained and atomistic potential are interpolated, and not the forces as in AdResS. Next, we impose that the Hamiltonian system must have the same thermodynamic equilibrium of the original force-based AdResS system; this is done by introducing a thermodynamic force in the auxiliary Hamiltonian approach, which, at the target temperature, keeps the density of particles across the system as in AdResS. \recheck{In the Hamiltonian approach we have the same equilibrium of the original adaptive (and full atomistic) system  and the difference between the work of the original thermodynamic force and the work of the thermodynamic force calculated in the Hamiltonian approach gives $\omega_{Q}$ (further details about this point are given in the Appendix)}. Moreover we have proven numerically, for the case of liquid water, that $\omega_{extra}=\int_{\Delta}\langle w\nabla_{x} w(x)( V_{AT}-V_{CG})\rangle_{\bf r} d{\bf r}$, where $V_{AT}$ and $V_{CG}$ are the atomistic and coarse-grained potential. 
\recheck{In the next section of this work we show analytically that the formula above is exact (at least) at the first order w.r.t. the probability distribution of the system as defined in Eq.\ref{order}}.
The result above implies that $\omega_{Q}$ can be calculated in a straightforward way during the initial equilibration within in the standard GC-AdResS code. 
\recheck{It must be noticed that, within the AdResS scheme, an approach similar to the auxilary Hamiltonian  has been recently proposed and applied to mixtures  by Potestio {\it et al.} \cite{h-adress} (see also \cite{luigientropy} where such an approach is commented)}. At this point according to \eqref{mu}, if one knows $\mu_{CG}$, then  GC-AdResS can automatically provide $\mu_{AT}$. However we need to do one step more, in fact the quantity of interest is not the total chemical potential, but the excess chemical potential $\mu^{ex}_{AT}$ which corresponds to the expression of \eqref{mu} where the kinetic (ideal gas) part is subtracted. Regarding the kinetic part, one can notice that the contribution coming from the center of mass is the same for the coarse-grained and for the atomistic molecules, thus it is automatically removed in the calculation of \eqref{mu}.
The kinetic part of $\mu_{AT}$ due to the rotational and vibrational degrees of freedom corresponds in our case to $\omega_{DOF}$ and \recheck{in principle can be calculated by hand (chemical potential of an isolated molecules). However such a calcualtion may become rather tedious for large and/or complex molecules but in our case it is actually not required.} In fact the technical set up of AdResS considers the removed degrees of freedom as phantom variables but thermally equilibrate them anyway \cite{simon-ch}. Thus the heat provided by the thermostat for the rotational and vibrational part is the same in the atomistic and coarse-graining molecules and is automatically removed in the difference. Finally, the calculation of $\mu^{ex}_{CG}$ can be done with standard methods, TI or IPM, which for simple spherical molecules, like those of the coarse-grained system, requires a negligible computational cost.
In conclusion, we have the final expression:
\mu^{ex}_{AT}=\mu^{ex}_{CG}-\int_{\Delta}{F}_{th}({\bf r})d{\bf r}-\int_{\Delta}\langle w\nabla_{x} w(x) (V_{AT}-V_{CG})\rangle_{\bf r} d{\bf r}
\label{finalmu}
\section*{Analytic eriviation of the $w_{extra}$}
\recheck{In this section we derive analytically the equivalence: $\omega_{extra}=\int_{\Delta}\langle w\nabla_{x} w(x)( V_{AT}-V_{CG})\rangle_{\bf r} d{\bf r}$ and define its limitations.\\
We define a potential coupling the atomistic and coarse-grained resolution as the spatial interpolation of the atomistic and coarse-grained potential, as done instead for the forces in the standard AdResS:
\begin{align}
  V =
  \sum_{i<j} w(\vect r_i) w(\vect r_j)
  V^\AT_{i,j}
  \sum_{i<j}[1- w(\vect r_i) w(\vect r_j)]
  V^\CG_{i,j},
where  $\vect r_i$ denotes
the center-of-mass (COM) position of the molecule with index $i$.
$V^\AT_{i,j}$ and $V^\CG_{i,j}$ are the atomistic and coarse-grained interaction potential between molecule $i$ and $j$, respectively, with:
\begin{align}
  V^\AT_{i,j} =
  \sum_{\alpha\in i}\sum_{\beta\in j} V^\AT(\vect r_\alpha - \vect r_\beta),\quad
  V^\CG_{i,j} =
  V^\CG(\vect r_i - \vect r_j).
where index $\alpha$ and $\beta$ denotes the atom position of the corresponding molecule.
The center of mass of the molecule is defined as:
\begin{align}
  \vect r_i = \sum_{\alpha\in i} \frac{m_\alpha}{\sum_{\alpha\in i} m_\alpha} \vect r_\alpha,
where $m_\alpha$ is the mass of atom $\alpha$.
The
corresponding intermolecular force is given by:
\begin{align}\label{eqn:interpol-f}
  \vect F_{i,j} =
  w(\vect r_i) w(\vect r_j)
  \vect F^\AT_{i,j}
  [1- w(\vect r_i) w(\vect r_j)]
  \vect F_{i,j}^\CG
  \nabla w(\vect r_i) w(\vect r_j)
  (V^\AT_{i,j} - V^\CG_{i,j})
We define the force of changing representation:
\begin{align}\label{eqn:frep}
  \vect F_{\res,i} = 
  \sum_j \nabla w(\vect r_i) w(\vect r_j)
  (V^\AT_{i,j} - V^\CG_{i,j})
%An important fact is that the force of changing representation is not
%symmetric w.r.t molecule $\alpha$ and $\beta$, therefore, the Newton's
%action-reaction law (momentum conservation) does not hold anymore for
%the potential interpolation approach. On the other hand, the force
%interpolation scheme does not preserve
%energy, while the potential interpolation
%approach has the advantage of energy conservation. In sum, both the
%force and potential interpolation approaches do not hold all physical
%conservation laws, because the hybrid region is artificially
%designed. Therefore, we will take the advantage of both and
%develop the thermodynamic relation (between the atomistic and
%coarse-grained resolutions)  for them at the same time.\\

We use the same notation as in our previous work~\cite{prx}.
The thermodynamic variables for the atomistic and coarse-grained regions
are denoted by $(N_1, V_1, T)$ and $(N_3, V_3, T)$, respectively.
We assume that the hybrid region is an infinitely thin filter (that is a much smaller region than the atomistic and coarse-grained region) that allows molecules
to change resolution as they cross it. Therefore, it is 
reasonable to assume that:
\begin{align}
  V &= V_1 + V_3\\
  N &= N_1 + N_3
\end{align}
where $V$ and $N$ are the total volume and total number of
molecules of the system. In this work, we adopt the same
assumptions as those listed in Sec.III.C of Ref.~\cite{prx},
i.e. we assume to be in the thermodynamic limit, and molecules are short-range
correlated (short-ranged must be intended as a range comparable to the size of the filter). The thermodynamic force for GC-AdResS ($\vect F_\thf$) and for the approach with the auxiliary Hamiltonian ($\vect F^\hadress_\thf$, i.e. potential interpolation AdResS),
enforce the system to have a flat density:
\begin{align}
  \rho_\HY &= \rho_\AT = \rho_\CG = \rho_0\\
  \rho_\HY^\hadress &= \rho_\AT = \rho_\CG = \rho_0
\end{align}
Where $\rho_0$ is the equilibrium number density of the system defined by $\rho_0 = N/V$.
\
As shown in Refs.\cite{prl12, prx} $\vect F_\thf$ provides the
balance of the grand potential or equivalently
\begin{align}\label{eqn:peq}
  p_\AT = p_\CG - \rho_0 \,\omega_\thf 
\end{align}
where $p$ denotes the pressure, and $ \omega_\thf$ denotes the
integral of the thermodynamic force.} \\

Instead when we consider the auxiliary Hamiltonian approach,  the force of changing representation is not
symmetric w.r.t molecule $\alpha$ and $\beta$, therefore, the Newton's

@@

Now assume, for simplicity, that the system changes resolution only along the $x$ direction.

@@

region.  The volume of the hybrid region is kept constant as if an ideal ``piston'' moves toward the CG region by an amount $\Delta L$.

@@

For infinitesimal displacement , $\Delta L$, we mean a displacement which is much smaller than the size of the
hybrid region. This is achievable by taking the limit of $\Delta L\rightarrow 0$,

@@

Therefore, the change of the free energy of the system is

@@

  \int d r\, \rho_0 S\cdot \Delta L\cdot
  \vect F^\hadress_\thf(r)

@@

  \int d r\, \rho_0 S\cdot \Delta L\cdot
  \big\langle \vect F_\res(r)\big\rangle

@@

where $A_\AT$ and $A_\CG$ is the free energy of the AT and CG region, respectively \recheck{HAN: why is it justified to split the free energy in free energies of subsystems and why the free energy of the HYB region is only the work and not also its kinetic component. I understand that since the molecule do not change resolution, their kinetic part stays the same and thus the difference is zero. However this formula needs either a reference or few more explanations}.
$N_1$ and $N_3$ is the number of molecules in the AT and CG region, and
$V_1$ and $V_3$ is the volume in the AT and CG region, respectively. $T$ is the temperature

@@

where $w_\thf^\hadress$ is the integral of the potential interpolation

>>>>

\def\pd#1#2{\frac{\partial #1}{\partial #2}}
\renewcommand{\vec}[1]{\mbox{\boldmath \small $#1$}}
\newcommand{\concen}{C}
\setlength\parindent{2em} 
\newcommand{\recheck}[1]{{\color{red} #1}}
\newcommand{\redc}[1]{{\color{red} #1}}            
\newcommand{\bluec}[1]{{\color{blue} #1}}            
\newcommand{\vect}[1]{\textbf{\textit{#1}}}
\newcommand{\dof}{{\textrm{DOF}}}
\newcommand{\AT}{{\textrm{{AT}}}}
\newcommand{\EX}{{\textrm{EX}}}
\newcommand{\CG}{{\textrm{CG}}}
\newcommand{\HY}{{\Delta}}
\newcommand{\rdf}{{\textrm{rdf}}}
\newcommand{\thf}{{\textrm{th}}}
\newcommand{\res}{{\textrm{rep}}}
\newcommand{\ext}{{\textrm{extra}}}
\newcommand{\exc}{{\textrm{ex}}}
\newcommand{\thermo}{{\textrm{Q}}}
\newcommand{\hadress}{{\textrm{H}}}
\newcommand{\dadress}{{\textrm{D}}}

@@

\bluec{
  Unified of terminology:
  \begin{itemize}
  \item interface region $\rightarrow$ transition region
  \item hybrid region $\rightarrow$ transition region
  \item define COM, change all ``center of mass'' to COM
  \item change the title of sec.III
  \item do not be too mean, we cite BOTH of their PRL H-Adress paper...
  \item some times, it is called GC-AdResS, sometimes original AdResS, sometimes standard AdResS, sometimes force-interpolation AdResS....
  \item rewrite eq. 4, see if it is OK for you
  \end{itemize}
}

\section{Introduction}
The chemical potential represents an important thermodynamic information for any system, in particular for liquids, where the possibility of combining different substances for forming optimal mixtures is strictly related to knowledge of the chemical potential of each component in the mixture environment. In this perspective, molecular simulation represents a powerful tool for predicting the chemical potential of complex molecular systems. Popular, well established methodologies in Molecular Dynamics (MD) are Widom particle insertion (IPM) \cite{widom} and thermodynamic integration (TI) \cite{ti}. IPM is computationally very demanding often beyond a reasonable limit even in presence of large computational resources, but upon convergence, is rather accurate. TI is computationally convenient but specifically designed to calculate the chemical potential and thus  it may not be optimal for employing MD for studying other properties. In fact TI requires artificial modification of the atomistic interactions (see Appendix).
Recently we have suggested that the chemical potential could be calculated by employing the Adaptive Resolution Simulation method in its Grand Canonical-like formulation (GC-AdResS) \cite{prl12,jctchan,prx}.
AdResS was originally designed  to interface regions of space at different levels of molecular resolution within one simulation set up. This allows for large and efficient multiscale simulations where the high resolution region is restricted to a small portion of space and the rest of the system is at coarser level. The recent version of the method, GC-AdResS, given its theoretical framework, should automatically calculate the chemical potential during the process of initial equilibration: in this work we prove that this is indeed the case and report results for the chemical potential for various liquids and mixtures of particular relevance in (bio)-chemistry and material science. We compare our results with those from full atomistic TI and find a satisfactory agreement. This agreement allows us to conclude that every time a multiscale GC-AdResS is performed, $\mu^{ex}$ is automatically calculated for each liquid component and implicitly confirm that the basic thermodynamics of the system is well described by the method.
\recheck{Moreover, in recent work AdResS has been merged with the MARTINI force field \cite{matej-sw1,matej-sw2}. In this context, the possibility of having a quantity like the chemical potential under \bluec{what is ``explicit control''} can be used as a further argument for the validity of the method in applications to large systems of biological interest.}
Below we provide the basic technical ingredients of GC-AdResS which are relevant for the calculation of the chemical potential, more specific details can be found in~\cite{jctchan, prx}.

@@

from the corresponding coarse-grained potential, \recheck{$\vect r$ is the center of mass (COM) position of}

@@

to $1$ (or vice versa) in the transition region ($\Delta$) where the lower resolution is then

@@

as illustrated in Fig.\ref{fig1}.
\item In the transition region a thermodynamic force acting on the COM of each molecule and a locally acting thermostat are added to assure the overall thermodynamic equilibrium at a given temperature. The thermodynamic force is defined in such a way that
  $p_{\AT}+\rho_{0}\int_{\Delta}{\vect F}_{th}({\vect r})d{\vect r}=p_{\CG}$,
  where $p_{AT}$  is the target pressure of the atomistic system (region), $p_{CG}$ is the pressure of the coarse-grained model, $\rho_{0}$ is the target molecular density of the atomistic system (region)~\cite{prl12}. An additional locally acting thermostat is added to take care of the loss/gain of energy in the transition region.

@@

$w(x)$ (in green. \recheck{Here the space is devided in $x$ direction, so it is a function of only $x$.}) is defined, and the region on the left, is the high resolution region (atomistic). It must be noticed that differently from the original AdResS, in GC-AdResS the range of definition of $w(x)$ is extended of an amount of ${\vect R}_{c}$. The extension of this additional region is equal to the cutoff radius of the atomistic interactions and $w(x)$ takes the constant value of $1$. The consequence is that molecules in the atomistic region interact with the rest of the system always and only via well defined atomistic interactions. This characteristic, in turn, allows to write an exact Hamiltonian for the atomistic region and thus treat the system in a Grand-Canonical fashion.\label{fig1}}

@@

In \cite{prx} we have defined necessary conditions in $\Delta$ such that the spatial probability distribution of the full-atomistic reference system was reproduced up to a certain (desired) order in the atomistic region of the adaptive system. \recheck{We have defined the $m$th order of a spatial (configurational) probability distribution of $N$ molecules, $p({\vect r}_{1}, \cdots, {\vect r}_{N})$, as:

@@

p^{(m)}({\vect r}_{1}, \cdots, {\vect r}_{m})=\int p({\vect r}_{1}, \cdots, {\vect r}_{m},{\vect r}_{m+1}, \cdots, {\vect r}_{N}) \: d{\vect r}_{m+1}\cdots d{\vect r}_{N}

@@

The first order, often mentioned in this work corresponds to the molecular number density $\rho({\vect r})$.}

@@

In the simulation set up, ${\vect F}_{th}$ is calculated via an iterative procedure \recheck{using} the molecular number density in $\Delta$. The iterative scheme consists of calculating \recheck{${\vect F}^{k+1}_{th}({\vect r}) = {\vect F}^{k}_{th}({\vect r}) + \frac{1}{\rho^2_{0}\kappa_{T}}\nabla \rho^{k}({\vect r})$} ($\kappa_{T}$ the isothermal compressibility), and the thermodynamic force is considered converged when the target density $\rho_{0}$ is reached in $\Delta$. As a result, ${\vect F}_{th}({\vect r})$, acting in $\Delta$, assures that there are no artificial density variations across the system, thus it allows to accurately reproduce the first order of the probability distribution in the atomistic region. Higher orders can be systematically achieved by imposing in $\Delta$ a corrective force.
% proportional to the gradient of the corresponding higher order of the probability distribution;
\recheck{For example, the COM-COM radial distribution function correction for the second order~\cite{jctchan}.} Next it was proved that indeed the target Grand Canonical distribution, that is the probability distribution of a subsystem (of the size of the atomistic region in GC-AdResS) in a large full atomistic simulation is accurately reproduced. A large number of tests were performed and the reproduction by GC-AdResS of the probability distribution was numerically proved up to (at least) the third order, more than sufficient in MD simulations. Within this framework it was finally shown that the sum of work of ${\vect F}_{th}({\vect r})$ and that of the thermostat corresponds to the difference in chemical potential between the atomistic and coarse-grained resolution; this subject is treated in the next section.

\section{Calculation of Excess Chemical Potential}

In Ref.~\cite{prx} it has been shown that the chemical potential of the atomistic and coarse-grained resolution are related by the following formula: 
\begin{equation}
\mu_{\CG}=\mu_{\AT}+\omega_{\thf}+\omega_{\thermo}
\label{mu}
\end{equation}
with $\mu_{\CG}$ the chemical potential of the coarse-grained system (in GC-AdResS this corresponds to a liquid of generic spheres), $\mu_{\AT}$  the chemical potential of the atomistic system, $\omega_{\thf}=\int_{\Delta}{\vect F}_{\thf}({\vect r})d{\vect r}$ the work of the thermodynamic force in the transition region, $\omega_{\thermo}$ the work/heat provided by the thermostat in order to slowly equilibrate the inserted/removed degrees of freedom in the transition region. $\omega_{\thermo}$ is composed by two parts, one, called $\omega_{\ext}$, which compensates the dissipation of energy due to the change of interactions in $\Delta$, and another, $\omega_{\dof}$, which is related to the equilibration of the reinserted/removed degrees of freedom (rotational and vibrational). {While the determination of $\omega_{\dof}$ is not required for the our final aim (that is the calcualtion of the excess chemical potential, as explain later on), the calculation of $\omega_{\thermo}$ is very relevant. However this calculation is not straightforward and we have proposed to introduce an auxiliary Hamiltonian approach where the coarse-grained and atomistic potential are interpolated, and not the forces as in AdResS. Next, we impose that the Hamiltonian system must have the same thermodynamic equilibrium of the original force-based AdResS system; this is done by introducing a thermodynamic force in the auxiliary Hamiltonian approach, which, at the target temperature, keeps the density of particles across the system as in AdResS. \recheck{In the auxiliary Hamiltonian approach we have the same equilibrium as the original adaptive (and full atomistic) system  and the difference between the work of the original thermodynamic force and the work of the thermodynamic force calculated in the Hamiltonian approach gives $\omega_{\ext}$ (further details about this point are given in the Appendix)}. Moreover we have proven numerically, for the case of liquid water, that
  $\omega_{\ext}=\int_{\Delta}\nabla w(\vect r)\langle w( U^{\AT}-U^{\CG})\rangle_{\vect r} d{\vect r}$,
  where $U_{AT}$ and $U_{CG}$ are the atomistic and coarse-grained potential. 
  \recheck{In the next section of this work we show analytically that the formula above is exact (at least) at the first order w.r.t. the probability distribution of the system as defined in Eq.\ref{order}}.

The result above implies that $\omega_{\thermo}$ can be calculated in a straightforward way during the initial equilibration within in the standard GC-AdResS code. 
\recheck{It must be noticed that, within the AdResS scheme, an approach similar to the auxiliary Hamiltonian  has been recently proposed and \bluec{applied to model pure and mixture systems} by Potestio {\it et al.} \cite{h-adress-0, h-adress} (see also \cite{luigientropy} where such an approach is commented)}. At this point according to \eqref{mu}, if one knows $\mu_{\CG}$, then  GC-AdResS can automatically provide $\mu_{\AT}$. However we need to do one step more, in fact the quantity of interest is not the total chemical potential, but the excess chemical potential $\mu^{\exc}_{\AT}$ which corresponds to the expression of \eqref{mu} where the kinetic (ideal gas) part is subtracted. Regarding the kinetic part, one can notice that the contribution coming from the COM is the same for the coarse-grained and for the atomistic molecules, thus it is automatically removed in the calculation of \eqref{mu}.
The kinetic part of $\mu_{\AT}$ due to the rotational and vibrational degrees of freedom corresponds in our case to $\omega_{\dof}$ and \recheck{in principle can be calculated by hand (chemical potential of an isolated molecules). However such a calcualtion may become rather tedious for large and/or complex molecules but in our case it is actually not required.} In fact the \recheck{Gromacs implementation of AdResS} considers the removed degrees of freedom as phantom variables but thermally equilibrate them anyway \cite{simon-ch}. Thus the heat provided by the thermostat for the rotational and vibrational part is the same in the atomistic and coarse-graining molecules and is automatically removed in the difference. Finally, the calculation of $\mu^{\exc}_{\CG}$ can be done with standard methods, TI or IPM, which for simple spherical molecules, like those of the coarse-grained system, requires a negligible computational cost.
In conclusion, we have the final expression:
\begin{equation}
  \mu^{\exc}_{\AT}=
  \mu^{\exc}_{\CG}-
  \int_{\Delta}{F}_{th}({\vect r})d{\vect r}-
  \int_{\Delta}\nabla_{\vect r} w(\vect r) \langle w(U^{\AT}-U^{\CG})\rangle_{\vect r} d{\vect r}
\label{finalmu}
\end{equation}

\section{Analytic Deriviation of the $w_{\ext}$}
{In this section we derive analytically the equivalence:
  $\omega_{\ext}=\int_{\Delta}\nabla_{\vect r} w(\vect r) \langle w(U^{\AT}-U^{\CG})\rangle_{\vect r} d{\vect r}$ and define its limitations.
We define a potential coupling the atomistic and coarse-grained resolution as the spatial interpolation of the atomistic and coarse-grained potential, as done instead for the forces in the standard AdResS:
\begin{align}
  U =
  \sum_{i<j} w(\vect r_i) w(\vect r_j)
  U^\AT_{i,j}
  +
  \sum_{i<j}[1- w(\vect r_i) w(\vect r_j)]
  U^\CG_{i,j},
\end{align}
where $U^\AT_{i,j}$ and $U^\CG_{i,j}$ are the atomistic and coarse-grained interaction potential between molecule $i$ and $j$, respectively, defined by
\begin{align}\label{eqn:pot-interpol}
  U^\AT_{i,j} =
  \sum_{\alpha\in i}\sum_{\beta\in j} U^\AT(\vect r_\alpha - \vect r_\beta),\quad
  U^\CG_{i,j} =
  U^\CG(\vect r_i - \vect r_j),
\end{align}
where index $\alpha$ and $\beta$ denotes the atom position of the corresponding molecule.
The COM of the molecule is defined as:
\begin{align}
  \vect r_i = \sum_{\alpha\in i} \frac{m_\alpha}{\sum_{\alpha\in i} m_\alpha} \vect r_\alpha,
\end{align}
where $m_\alpha$ is the mass of atom $\alpha$.
\recheck{The potential interpolation~\eqref{eqn:pot-interpol} provides an auxiliary Hamiltonian to the AdResS system, and
the
corresponding intermolecular force is given by:}
\begin{align}\label{eqn:interpol-f}
  \vect F_{i,j} =
  w(\vect r_i) w(\vect r_j)
  \vect F^\AT_{i,j}
  +
  [1- w(\vect r_i) w(\vect r_j)]
  \vect F_{i,j}^\CG
  -
  \nabla_{\vect r} w(\vect r_i) w(\vect r_j)
  (U^\AT_{i,j} - U^\CG_{i,j}).
\end{align}
\recheck{We refer to the AdResS simulation using force scheme~\eqref{eqn:interpol-f} as
  auxiliary Hamiltonian AdResS, and all properties of this approach will be added a superscript ``H''}.
We define the force of changing representation by
\begin{align}\label{eqn:frep}
  \vect F_{\res,i} = 
  \sum_j \nabla_{\vect r} w(\vect r_i) w(\vect r_j)
  (U^\AT_{i,j} - U^\CG_{i,j}).
\end{align}
%An important fact is that the force of changing representation is not
%symmetric w.r.t molecule $\alpha$ and $\beta$, therefore, the Newton's
%action-reaction law (momentum conservation) does not hold anymore for
%the potential interpolation approach. On the other hand, the force
%interpolation scheme does not preserve
%energy, while the potential interpolation
%approach has the advantage of energy conservation. In sum, both the
%force and potential interpolation approaches do not hold all physical
%conservation laws, because the hybrid region is artificially
%designed. Therefore, we will take the advantage of both and
%develop the thermodynamic relation (between the atomistic and
%coarse-grained resolutions)  for them at the same time.\\

We use the same notation as in our previous work~\cite{prx}.
The thermodynamic variables for the atomistic and coarse-grained regions
are denoted by $(N_1, V_1, T)$ and $(N_3, V_3, T)$, respectively.
We assume that the transition region is an infinitely thin filter (that is a much smaller region than the atomistic and coarse-grained region) that allows molecules
to change resolution as they cross it. Therefore, it is 
reasonable to assume that:
\begin{align}
  V &= V_1 + V_3\\
  N &= N_1 + N_3
\end{align}
where $V$ and $N$ are the total volume and total number of
molecules of the system. In this work, we adopt the same
assumptions as those listed in Sec.III.C of Ref.~\cite{prx},
i.e. we assume the system to be in the thermodynamic limit, and molecules are short-range
correlated (short-ranged must be intended as a range comparable to the size of the filter). The thermodynamic force for GC-AdResS ($\vect F_\thf$) \recheck{and for the auxiliary Hamiltonian AdResS ($\vect F^\hadress_\thf$),}
enforce the system to have a flat density:
\begin{align}
  \rho_\HY &= \rho_\AT = \rho_\CG = \rho_0\\
  \rho_\HY^\hadress &= \rho_\AT = \rho_\CG = \rho_0
\end{align}
Where $\rho_0$ is the equilibrium number density of the system defined by $\rho_0 = N/V$.
As shown in Refs.\cite{prl12, prx}, $\vect F_\thf$ provides the
balance of the grand potential or equivalently
\begin{align}\label{eqn:peq}
  p_\AT = p_\CG - \rho_0 \,\omega_\thf 
\end{align}
where $ \omega_\thf$ denotes the integral of the thermodynamic force $\vect F_\thf$. \\

Instead when we consider the auxiliary Hamiltonian approach, the third part of the R.H.S.~of Eq.~\eqref{eqn:interpol-f} is not
symmetric w.r.t molecule $i$ and $j$, therefore, the Newton's

@@

Now assume, \recheck{for simplicity and without lost of generality}, that the system changes resolution only along the $x$ direction.

@@

region.  The volume of the transition region is kept constant as if an ideal ``piston'' moves toward the CG region by an amount $\Delta L$.

@@

The displacement $\Delta L$ should be infinitesimal, i.e.~much smaller than the size of the
transition region. This is achievable by taking the limit of $\Delta L\rightarrow 0$,

@@

Therefore, the change of the free energy of the system is \recheck{approximately}:

@@

  \int_\Delta d \vect r\, \rho_0 S\cdot \Delta L\cdot
  \big[\vect F^\hadress_\thf(\vect r)

@@

  \big\langle \vect F_\res(\vect r)\big\rangle\,\big]

@@

where $A_\AT$ and $A_\CG$ are the free energies of the AT and CG region, respectively.
\recheck{Since the system is in the thermodynamic limit, and the interactions in the system are short-ranged,
  the total free energy has the above splitting form.}
% \recheck{HAN: why is it justified to split the free energy in free energies of subsystems and why the free energy of the HYB region is only the work and not also its kinetic component. I understand that since the molecule do not change resolution, their kinetic part stays the same and thus the difference is zero. However this formula needs either a reference or few more explanations}.
$N_1$ and $N_3$ are the numbers of molecules in the AT and CG region, and
$V_1$ and $V_3$ are the volumes in the AT and CG region, respectively. $T$ is the temperature

@@

where $w_\thf^\hadress$ is the integral of the 
